---
title: "Data Wrangling: R Style"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: zenburn
    center: true
    transition: fade
    css: clean.css
    slide_level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
library(dplyr)

```


## R

R is more than just a "stats package".

R is an object-oriented programming language.

A common question is, "Can R *insert something trivial*?".

## Is R Really Different?

* Cost

    * You cannot beat free

* Power

    * Object-oriented programming language with near-infinite flexability

* Open

    * Anyone can contribute to it and/or look at function code

* Maintainers/Package Creators

    * The people who develop methods create packages for those methods
    
  
## Installing packages

```{r install, eval = FALSE, echo = TRUE}
install.packages('tidyverse')
```

## The Basics

Everything in R is an object.

You must create an object and you can then call on the object.

- Always be sure to name objects something other than function names!

```{r , eval = TRUE, echo = TRUE}

numList = 1:5

numList
```

## Functions

Functions are operations on an object. 

```{r, eval = TRUE}
nonsenseFunction = function(x) {
  res = (sqrt(x) + sin(x))^2
  return(res)
}

test = 1:10

nonsenseFunction(test)
```


## Ready?

Now that you know we are just dealing with objects and functions, you are 90% of the way to R enlightenment.


## {data-background="tidyverseLogo.png"}


## The tidyverse

The tidyverse is an ecosystem for making data wrangling easy.

<span class="pack">dplyr</span>

<span class="pack">tidyr</span>

<span class="pack">stringr</span>

<span class="pack">magrittr</span>

<span class="pack">rvest</span>

<span class="pack">ggplot2</span>


## Riffing With dplyr

The <span class="pack">dplyr</span> package was created to make data manipulation easier.

It provides a sensible and coherent grammar for working with data.


## From this...

```{r baseRexample1, eval=FALSE, echo = TRUE}

newData = oldData[oldData$Z == 'Yes' & oldData$Q == 'No', ]
cols = c('ID', paste0('X', 1:10), 'var1', 'var2', 
         grep("^Merc[0-9]+", colnames(oldData), value = TRUE))
newData = newData[, cols]
newData = tail(newData, 50)
newData = sort(newData, decreasing = TRUE)

```

## To this...

```{r, eval = FALSE, echo = TRUE}
newData = oldData %>% 
  filter(Z == 'Yes', Q == 'No') %>% 
  select(num_range('X', 1:10), contains('var'), starts_with('Merc')) %>% 
  tail(50) %>% 
  arrange(desc(var1))
```


## A Quick Word on %>%

In the previous snippet, you hopefully noticed the <span class="func">\%\>\%</span>.

It is included in <span class="pack">dplyr</span>, but it originates in <span class="pack">magrittr</span>.

It is pronounced as <span class="emph">pipe</span> and is functionally equivalent to the Unix <span class="func">\|</span>

  - Check out <span class="pack">magrittr</span> for other pipes.


## Why?

Old-school R:

```{r, echo = TRUE, eval = FALSE}

ceiling(mean(abs(sample(-100:100, 50))))

```

Piping:

```{r, echo = TRUE, eval = FALSE}

-100:100 %>% 
  sample(50) %>% 
  abs %>% 
  mean %>% 
  ceiling

```


Both are valid, but one is just a bit easier for human eyes and easier to code.


## Our Data

<a href="https://en.wikipedia.org/wiki/List_of_highest-grossing_films">Highest Grossing Films</a>

```{r}
library(rvest); library(xml2); library(magrittr)

allTables = read_html("https://en.wikipedia.org/wiki/List_of_highest-grossing_films") %>% 
  html_table(fill = TRUE)

highestGrossing = allTables %>% 
  extract2(1)

head(highestGrossing)
```


## Congratulations...

You know how to scrape data from the web!

## Our Path Forward

Select variables

Clean and convert strings

Add new variables

Summarize

## Our Destination


## Select

The <span class="func">select</span> function does exactly what it says: it selects variables.

It also has some handy helpers for our convenience:

starts_with

ends_with

contains

matches

## Action!

```{r}
highestGrossing = highestGrossing %>% 
  select(-Rank, -`Reference(s)`)

head(highestGrossing)
```

## String Cleaning

Strings happen...

```{r}
library(stringr)

highestGrossing = highestGrossing %>% 
  mutate(Peak = str_replace(Peak, "[A-Z].*", ""), 
         Peak = as.numeric(Peak), 
         gross = as.numeric(str_replace_all(`Worldwide gross`, "\\$|,", ""))) %>% 
  select(-`Worldwide gross`)
```


## Filter

```{r}
highestGrossing = highestGrossing %>% 
  filter(Peak != 1 & Peak != max(Peak))
```


## Mutate...

#### Momentarily

Our dollar values are nominal, but I think we should add a real dollar column.

First, we will get a GDP deflator table and value:

```{r}
gdpDeflator2017 = read_html("http://www.multpl.com/gdp-deflator/table") %>% 
  html_table(trim = TRUE, header = TRUE) %>% 
  extract2(1) %>% 
  filter(grepl("2017", Date) == TRUE) %>% 
  select(-Date) %>% 
  extract2(1)
```

```{r}
gdpDeflator2017
```

## Using Mutate

Coherent grammar, huh? Would someone like to explain why <span class="func">mutate</span>?

```{r}
highestGrossing = highestGrossing %>% 
  mutate(grossReal2017 = gross/(gdpDeflator2017/100))
```

## Group_by

```{r}
highestGrossing = highestGrossing %>% 
  group_by(Year)
```

## Summarize

```{r}
highestGrossingSummary = highestGrossing %>% 
  summarize(n = n(), 
            meanGross = mean(grossReal2017))

highestGrossingSummary
```

## Visual Explorations

```{r}
library(ggplot2)

ggplot(highestGrossingSummary, aes(Year, meanGross, size = n)) +
  geom_point() +
  theme_minimal()
```


## A Little More Modern

```{r}
movieByYear = highestGrossing %>% 
  group_by(Year) %>%
  dplyr::summarise(movies = paste(Title, collapse = ", ")) %>% 
  right_join(., highestGrossingSummary, by = "Year")
```


```{r}
library(plotly)

p = ggplot(movieByYear, aes(Year, meanGross, size = n)) +
  geom_point(aes(text = movies)) +
  theme_minimal()

ggplotly(p, tooltip = c("text"))

```

# Other Cool dplyr Functions

## Coalesce

```{r}
testDF = data.frame(x = c(1:5, rep(NA, 5)), 
                    y = c(rep(NA, 5), 6:10))

testDF %>% 
  mutate(z = coalesce(x, y))
```





## Distinct












## Just Scratching The Surface

We have only really seen the tip of the iceberg with regard to what R has to offer.

- We did not even talk about all its analytical capabilities -- just know that it does anything!


```{r, eval = FALSE, echo = FALSE}
fortunes::fortune(which = 323)
```



## An Example

```{r, eval = FALSE, echo = c(-1)}
load("N:/R/conflictData/data/uppsalaConflictData.RData")
library(dygraphs); library(tidyr)
as.data.frame(groupTotals) %>%
  select(year, perp, tot) %>%
  group_by(year, perp) %>%
  summarize(tot = sum(tot)) %>%
  arrange(year) %>%
  spread(perp, tot) %>%
  dygraph() %>%
  dyLegend(show = "onmouseover") %>%
  dyHighlight(highlightSeriesBackgroundAlpha = .2,
              hideOnMouseOut = FALSE) %>%
  dyCSS("N:/R/conflictData/dygraphLegend.css")
```


## An Example {.flexbox .vcenter}

```{r, echo = FALSE, eval = FALSE, warning=FALSE}
load("conflictDat.RData")
library(dygraphs); library(tidyr)
as.data.frame(groupTotals) %>%
  select(year, perp, tot) %>%
  group_by(year, perp) %>%
  summarize(tot = sum(tot)) %>%
  arrange(year) %>%
  spread(perp, tot) %>%
  dygraph() %>%
  dyLegend(show = "onmouseover") %>%
  dyHighlight(highlightSeriesBackgroundAlpha = .2,
              hideOnMouseOut = FALSE) %>%
  dyCSS("dygraphLegend.css")
```


## Another Example (Because They Are Fun!)

```{r, warning = FALSE, message = FALSE, echo = TRUE, eval = FALSE}
library(plotly)

plot_ly(economics, x = date, y = uempmed) %>%
  add_trace(y = fitted(loess(uempmed ~ as.numeric(date))), x = date) %>%
  layout(title = "Median duration of unemployment (in weeks)",
         showlegend = FALSE) %>%
  dplyr::filter(uempmed == max(uempmed)) %>%
  layout(annotations = list(x = date, y = uempmed, text = "Peak", showarrow = T))
```


## Another Example (Because They Are Fun!)

```{r, warning = FALSE, message = FALSE, echo = FALSE, eval = FALSE}
library(plotly)

plot_ly(economics, x = date, y = uempmed) %>%
  add_trace(y = fitted(loess(uempmed ~ as.numeric(date))), x = date) %>%
  layout(title = "Median duration of unemployment (in weeks)",
         showlegend = FALSE) %>%
  dplyr::filter(uempmed == max(uempmed)) %>%
  layout(annotations = list(x = date, y = uempmed, text = "Peak", showarrow = T))
```


## Last One...

```{r, warning = FALSE, message = FALSE, echo = TRUE, eval = FALSE}
library(ggplot2)

p = ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(text = paste("Transmission:", as.factor(am))), size = 2) +
  geom_smooth(aes(colour = as.ordered(cyl), fill = as.ordered(cyl)), 
              show.legend = FALSE) + 
  facet_grid(. ~ cyl) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  #scale_colour_discrete(name = "Cylinders") +
  lazerhawk::theme_trueMinimal()

ggplotly(p)
```



## Last One...

```{r, warning = FALSE, message = FALSE, echo = FALSE, eval = FALSE}
p = ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(text = paste("Transmission:", as.factor(am))), size = 2) +
  geom_smooth(aes(colour = as.ordered(cyl), fill = as.ordered(cyl)), show.legend = FALSE) + 
  facet_grid(. ~ cyl) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  #scale_colour_discrete(name = "Cylinders") +
  lazerhawk::theme_trueMinimal()

ggplotly(p)
```



## Interactive Tables

```{r, echo = 3, eval = FALSE}
mtcars1 = mtcars
mtcars1$am = factor(mtcars1$am, levels = 0:1)
DT::datatable(head(mtcars1), filter = "top")
```




# Wrap

## Wrapping It Up

Although our time was brief, I hope you can see the benefits of using R/RStudio for your research.

R only continues to develop, improve, and grow!

By using R, you will also find growth!

  - Maybe not happiness, but definitely growth.
  
```{r, eval = FALSE, echo = FALSE}
fortunes::fortune(which = 83)
```


