---
title: "Data Wrangling: R Style"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: zenburn
    center: true
    transition: fade
    css: clean.css
    slide_level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE)
library(dplyr)
```


## R

R is more than just a "stats package".

- It is a statistical programming language!

R is an object-oriented programming language.

A common question is, "Can R do *this*?".

- The answer is always, "Yes, but can *you* do it?".

    - Just to give you an idea, <span class="pack">tuneR</span> can be used to make electronic music

## Is R Really Different?

* Cost

    * You cannot beat free

* Power

    * Object-oriented programming language with near-infinite flexability

* Open

    * Anyone can contribute to it and/or look at function code

* Maintainers/Package Creators

    * The people who develop methods create packages for those methods
    
  
## Installing packages

```{r install, eval = FALSE, echo = TRUE}
install.packages('tidyverse)
```

## {data-background="tidyverseLogo.png"}


## The Basics

Everything in R is an object.

You must create an object and you can then call on the object.

- Always be sure to name objects something other than function names!

```{r , eval = FALSE, echo = TRUE}
### <b>
numList = 1:5
###</b>
numList
```


## dplyr

The <span class="pack">dplyr</span> package was created to make data manipulation easier.


## From this...

```{r baseRexample1, eval=FALSE, echo = TRUE}
# two step with grep (searching with regular expressions)

newData = oldData[oldData$Z == 'Yes' & oldData$Q == 'No', ]

cols = c('ID', paste0('X', 1:10), 'var1', 'var2', 
         grep("^Merc[0-9]+", colnames(oldData), value = TRUE))

newData = newData[, cols]

newData = tail(newData, 50)

newData = sort(newData, decreasing = TRUE)

```

## To this...

```{r, eval = FALSE, echo = TRUE}
newData = oldData %>% 
  filter(Z == 'Yes', Q == 'No') %>% 
  select(num_range('X', 1:10), contains('var'), starts_with('Merc')) %>% 
  tail(50) %>% 
  arrange(desc(var1))
```



## Select

starts_with

ends_with

contains

matches

## Filter

## Mutate

## Coalesce

## Group_by

## Distinct

## Summarize









## A Quick Word on %>%

In the previous snippet, you hopefully noticed the <span class="func">\%\>\%</span>.

It is included in <span class="pack">dplyr</span>, but it originates in <span class="pack">magrittr</span>.

It is pronounced as <span class="emph">pipe</span> and is functionally equivalent to the Unix <span class="func">\|</span>

  - Check out <span class="pack">magrittr</span> for other pipes.


## Why?

Old-school R:

```{r, echo = TRUE, eval = FALSE}

ceiling(mean(abs(sample(-100:100, 50))))

```

Piping:

```{r, echo = TRUE, eval = FALSE}

-100:100 %>% 
  sample(50) %>% 
  abs %>% 
  mean %>% 
  ceiling

```


Both are valid, but one is just a bit easier for human eyes and easier to code.

## Just Scratching The Surface

We have only really seen the tip of the iceberg with regard to what R has to offer.

- We did not even talk about all its analytical capabilities -- just know that it does anything!


```{r, eval = FALSE, echo = FALSE}
fortunes::fortune(which = 323)
```



## An Example

```{r, eval = FALSE, echo = c(-1)}
load("N:/R/conflictData/data/uppsalaConflictData.RData")
library(dygraphs); library(tidyr)
as.data.frame(groupTotals) %>%
  select(year, perp, tot) %>%
  group_by(year, perp) %>%
  summarize(tot = sum(tot)) %>%
  arrange(year) %>%
  spread(perp, tot) %>%
  dygraph() %>%
  dyLegend(show = "onmouseover") %>%
  dyHighlight(highlightSeriesBackgroundAlpha = .2,
              hideOnMouseOut = FALSE) %>%
  dyCSS("N:/R/conflictData/dygraphLegend.css")
```


## An Example {.flexbox .vcenter}

```{r, echo = FALSE, eval = FALSE, warning=FALSE}
load("conflictDat.RData")
library(dygraphs); library(tidyr)
as.data.frame(groupTotals) %>%
  select(year, perp, tot) %>%
  group_by(year, perp) %>%
  summarize(tot = sum(tot)) %>%
  arrange(year) %>%
  spread(perp, tot) %>%
  dygraph() %>%
  dyLegend(show = "onmouseover") %>%
  dyHighlight(highlightSeriesBackgroundAlpha = .2,
              hideOnMouseOut = FALSE) %>%
  dyCSS("dygraphLegend.css")
```


## Another Example (Because They Are Fun!)

```{r, warning = FALSE, message = FALSE, echo = TRUE, eval = FALSE}
library(plotly)

plot_ly(economics, x = date, y = uempmed) %>%
  add_trace(y = fitted(loess(uempmed ~ as.numeric(date))), x = date) %>%
  layout(title = "Median duration of unemployment (in weeks)",
         showlegend = FALSE) %>%
  dplyr::filter(uempmed == max(uempmed)) %>%
  layout(annotations = list(x = date, y = uempmed, text = "Peak", showarrow = T))
```


## Another Example (Because They Are Fun!)

```{r, warning = FALSE, message = FALSE, echo = FALSE, eval = FALSE}
library(plotly)

plot_ly(economics, x = date, y = uempmed) %>%
  add_trace(y = fitted(loess(uempmed ~ as.numeric(date))), x = date) %>%
  layout(title = "Median duration of unemployment (in weeks)",
         showlegend = FALSE) %>%
  dplyr::filter(uempmed == max(uempmed)) %>%
  layout(annotations = list(x = date, y = uempmed, text = "Peak", showarrow = T))
```


## Last One...

```{r, warning = FALSE, message = FALSE, echo = TRUE, eval = FALSE}
library(ggplot2)

p = ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(text = paste("Transmission:", as.factor(am))), size = 2) +
  geom_smooth(aes(colour = as.ordered(cyl), fill = as.ordered(cyl)), 
              show.legend = FALSE) + 
  facet_grid(. ~ cyl) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  #scale_colour_discrete(name = "Cylinders") +
  lazerhawk::theme_trueMinimal()

ggplotly(p)
```



## Last One...

```{r, warning = FALSE, message = FALSE, echo = FALSE, eval = FALSE}
p = ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(text = paste("Transmission:", as.factor(am))), size = 2) +
  geom_smooth(aes(colour = as.ordered(cyl), fill = as.ordered(cyl)), show.legend = FALSE) + 
  facet_grid(. ~ cyl) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  #scale_colour_discrete(name = "Cylinders") +
  lazerhawk::theme_trueMinimal()

ggplotly(p)
```



## Interactive Tables

```{r, echo = 3, eval = FALSE}
mtcars1 = mtcars
mtcars1$am = factor(mtcars1$am, levels = 0:1)
DT::datatable(head(mtcars1), filter = "top")
```




# Wrap

## Wrapping It Up

Although our time was brief, I hope you can see the benefits of using R/RStudio for your research.

R only continues to develop, improve, and grow!

By using R, you will also find growth!

  - Maybe not happiness, but definitely growth.
  
```{r, eval = FALSE, echo = FALSE}
fortunes::fortune(which = 83)
```


