---
title: "Introduction to R/RStudio"
author: "Seth Berry"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: zenburn
    center: true
    transition: fade
    css: clean.css
    slide_level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
library(dplyr)
```


## R

R is more than just a "stats package".

- It is a statistical programming language!

R is an object-oriented programming language.

A common question is, "Can R do *this*?".

- The answer is always, "Yes, but can *you* do it?".

    - Just to give you an idea, <span class="pack">tuneR</span> can be used to make electronic music

## Is R Really Different?

* Cost

    * You cannot beat free

* Power

    * Object-oriented programming language with near-infinite flexability

* Open

    * Anyone can contribute to it and/or look at function code

* Maintainers/Package Creators

    * The people who develop methods create packages for those methods
    
    
## Common Hang-ups

"SAS/SPSS/Stata can do chi-square, is there really a difference?"

"I am not a programmer."

"R is really hard/R has a huge learning curve."
    
"I don't have time to learn."

```{r, eval = TRUE, echo = FALSE}
fortunes::fortune(which = 52)
```

- use the <span class="pack">fortunes</span> package for inspiration!    


## Relevant

```{r, echo = FALSE, eval = TRUE}
fortunes::fortune(which = 108)
```
    


## Is It The Only One?

Of course not!

Despite being able to do absolutely anything, somethings are easier to do outside of R.

- Just remember that you need to try to make your work reproducible.

    - Just don't do anything by hand!

You can use any number of packages for your research, but R is very close to a "one-stop-shop".



## CRAN

The Comprehensive R Archive Network is the "official" package repository for R.

> - There are currently <span class="emph">9044</span> packages on CRAN.

> - For perspective, R has 200X more functions than SAS.
  
  
## Finding Information

CRAN Task Views allow you to see a variety of functions associated with topics.

| Task View Examples | Example Packages |
|-----------+------------------|
| Econometrics | <span class="pack">wbstats</span> & <span class="pack">plm</span> |
| Finance | <span class="pack">quantmod</span> & <span class="pack">urca</span> |
| Machine Learning | <span class="pack">rpart</span> & <span class="pack">caret</span> |
| Natural Language Processing | <span class="pack">tm</span> & <span class="pack">koRpus</span> |
| Psychometrics | <span class="pack">lavaan</span> & <span class="pack">mirt</span> |
| Spatial | <span class="pack">sp</span> & <span class="pack">rgdal</span> |
| Time Series | <span class="pack">zoo</span> & <span class="pack">forecast</span> |

  
## Installing packages

From CRAN

```{r install, eval = FALSE, echo = TRUE}
install.packages(c("devtools", "dplyr"))
```

From GitHub:

```{r githubInstall, eval = FALSE, echo = TRUE}
devtools::install_github("saberry/qualtricsR")
```

- The <span class="pack">devtools</span> package also has the ability to install packages from other repositories (e.g., bitbucket, svn).

## The Basics

Everything in R is an object.

You must create an object and you can then call on the object.

- Always be sure to name objects something other than function names!

```{r , eval = TRUE, echo = TRUE}
### <b>
numList = 1:5
###</b>
numList
```

- <span class="func">\<-</span> is the classic assignment operator.

```{r, eval = TRUE, echo = TRUE}
numList * 5
```


## Object Types

R has many different kinds of objects:

Item

- Numeric

- Character

- Factor/ordered

Data

- Data frame

- Matrix

- List


## The Index

Because R creates objects, each object can be manipulated through an index.

Like many other languages, an object's index is generally accessed using <span class="func">\[\]</span>:

```{r index, eval = TRUE, echo = TRUE}
### <b>
numList[1:3]
###</b>
```

```{r indexMult, eval = TRUE, echo = TRUE}
numList[1:3] * 5
```

For *named* objects, we can use the <span class="func">$</span>:

```{r monIndex, eval = TRUE, echo = TRUE}
head(mtcars$mpg)
```

## A Big Index Hint

Just like matrix algebra and dimensional lumber -- obj[*rows*, *columns*]

```{r, eval = TRUE, echo = TRUE}
mtcars[1, ]
```

```{r, eval = TRUE, echo = TRUE}
head(mtcars[, 1])
```

```{r, eval = TRUE, echo = TRUE}
mtcars[1, 1]
```



## Operators and Math Functions

Like any other language (or program, for that matter), R has the ability to use operators:

```{r ops, eval = TRUE, echo = TRUE}
mtcars$mpg[mtcars$cyl == 6 | mtcars$cyl == 8 & mtcars$hp >= 146]
```

And math functions:

```{r, eval = TRUE, echo = TRUE}
sqrt((2 + 2)^2 * (7 / (2 - 1))) * pi
```


## Basic Functions

Even with all of the packages that R has, base R is still extremely powerful by itself.

```{r strEx, eval = TRUE, echo = TRUE}
str(numList)
```

```{r summaryEx, eval = TRUE, echo = TRUE}
summary(numList)
```

```{r meanEx, eval = TRUE, echo = TRUE}
mean(numList)
```


## Basic Analyses

```{r, eval = TRUE, echo = TRUE}
cor(mtcars$mpg, mtcars$wt)
```


```{r, eval = TRUE, echo = TRUE}
lm(mpg ~ wt, data = mtcars)
```



## Basic Plotting

```{r, echo = TRUE}
plot(mtcars$wt, mtcars$mpg, pch = 19)
```


## Combining Functions

R allows you to combine functions:

```{r, eval = TRUE, fig.height=4, echo = TRUE}
plot(mtcars$wt, mtcars$mpg, pch = 19)
lines(lowess(mtcars$wt, mtcars$mpg), col = "#FF6600", lwd = 2)
abline(lm(mpg ~ wt, data = mtcars), col = "#0099ff", lwd = 2)
```

## Modern Approaches For Data Wrangling

We saw a glimpse of what base R has to offer in terms of data manipulation.

As powerful as the indexing approach may be, it can often be messy and slightly confusing to someone who may be interested in using your code (or the future you).

- Because it is object-oriented, it is inherently more powerful than traditional stats programs.

## An Example

```{r baseRexample1, eval=FALSE, echo = TRUE}
### NICE R DATA ###

# numeric indexes; not conducive to readibility or reproducibility
newData = mtcars[, 1:4]

# explicitly by name; fine if only a handful; not pretty
newData = mtcars[, c('mpg','cyl', 'disp', 'hp')]

### MEAN REAL DATA ###

# two step with grep (searching with regular expressions)
cols = c('ID', paste0('X', 1:10), 'var1', 'var2', 
         grep("^Merc[0-9]+", colnames(oldData), value = TRUE))

newData = oldData[, cols]

# or via subset
newData = subset(oldData, select = cols)
```


## More
What if you also want observations where **Z** is **Yes**, **Q** is **No**, and only the last 50 of those results, ordered by **var1** (descending)?

```{r baseRexample2, eval=FALSE, echo = TRUE}
# three operations and overwriting or creating new objects if we want clarity
newData = newData[oldData$Z == 'Yes' & oldData$Q == 'No', ]
newData = tail(newData, 50)
newData = newdata[order(newdata$var1, decreasing = TRUE), ]
```

And this is for fairly straightforward operations.

## dplyr

The <span class="pack">dplyr</span> package was created to make data manipulation easier.

```{r, eval = FALSE, echo = TRUE}
newData = oldData %>% 
  filter(Z == 'Yes', Q == 'No') %>% 
  select(num_range('X', 1:10), contains('var'), starts_with('Merc')) %>% 
  tail(50) %>% 
  arrange(desc(var1))
```


## Other Handy Functions

```{r, eval = TRUE, echo = TRUE}
mtcars %>% 
  filter(am == 0) %>% # Automatic transmission
  select(mpg, cyl, hp, wt) %>% 
  mutate(rawWeight = wt * 1000) %>% 
  group_by(cyl) %>% 
  summarize_all(funs(mean)) 
```


## coalesce

```{r, eval = TRUE, echo = TRUE}

x = c(1, 2, NA, NA, 5, 6, NA, 8, NA, NA)

y = c(NA, NA, 3, 4, NA, NA, NA, NA, NA, NA)

z = c(NA, NA, NA, NA, NA, NA, 7, NA, 9, 10 )

coalesce(x, y, z)

```

## A Quick Word on %>%

In the previous snippet, you hopefully noticed the <span class="func">\%\>\%</span>.

It is included in <span class="pack">dplyr</span>, but it originates in <span class="pack">magrittr</span>.

It is pronounced as <span class="emph">pipe</span> and is functionally equivalent to the Unix <span class="func">\|</span>

  - Check out <span class="pack">magrittr</span> for other pipes.


## Why?

Old-school R:

```{r, echo = TRUE, eval = FALSE}

ceiling(mean(abs(sample(-100:100, 50))))

```

Piping:

```{r, echo = TRUE, eval = FALSE}

-100:100 %>% 
  sample(50) %>% 
  abs %>% 
  mean %>% 
  ceiling

```


Both are valid, but one is just a bit easier for human eyes and easier to code.

## Just Scratching The Surface

We have only really seen the tip of the iceberg with regard to what R has to offer.

- We did not even talk about all its analytical capabilities -- just know that it does anything!

Do take some time to look through the CRAN Task Views.

The RBloggers website always has new and neat stuff.

Daily and weekly trending repositories on GitHub are also enlightening.

```{r, eval = TRUE, echo = FALSE}
fortunes::fortune(which = 323)
```


## RStudio

> - RStudio is an integrated development environment (IDE) for R.

> - It was developed by J.J. Allaire and Hadley Wickham is the Chief Scientist.

> - There are others, but all pale in comparison.

## Scripting

Before RStudio, there were other options beside using the console.

- My heart goes out to everyone else who has learned on the console and/or Tinn-R.

In scripting with RStudio, you are getting:

    - Code completion (use tab to autocomplete anything)
    - Code highlighing
    - Code diagnostics/warnings
    - Code snippets (tab for apply and loops)
    - Easily accessible help files (F1 on any function)
    - Code tidying (Ctrl + Shift + A)
    - More shortcuts than you can learn (Alt + Shift + K)
    - Automatic pairing of closures (or...ruining your typing)

## File Types

In addition to R scripts, RStudio offers an array of file types:

* Text
    + JS
    + CSS
    + Python
* Markdown
* Presentation
* C++

Any language file has the code highlighted and diagnosed.

## Interactive Data Exploration

With an assortment of the <span class="objclass">htmlwidgets</span> packages, we can create a wide variety of output.

- <span class="pack">ggvis</span>, <span class="pack">rbokeh</span>, <span class="pack">plotly</span>: interactive visualizations
- <span class="pack">dygraphs</span>: interactive time-series visualizations
- <span class="pack">DT</span>: interactive tables
- <span class="pack">Leaflet</span>: geo-spatial maps

## An Example

```{r, eval = FALSE, echo = c(-1)}
load("N:/R/conflictData/data/uppsalaConflictData.RData")
library(dygraphs); library(tidyr)
as.data.frame(groupTotals) %>%
  select(year, perp, tot) %>%
  group_by(year, perp) %>%
  summarize(tot = sum(tot)) %>%
  arrange(year) %>%
  spread(perp, tot) %>%
  dygraph() %>%
  dyLegend(show = "onmouseover") %>%
  dyHighlight(highlightSeriesBackgroundAlpha = .2,
              hideOnMouseOut = FALSE) %>%
  dyCSS("N:/R/conflictData/dygraphLegend.css")
```


## An Example {.flexbox .vcenter}

```{r, echo = FALSE, eval = TRUE, warning=FALSE}
load("conflictDat.RData")
library(dygraphs); library(tidyr)
as.data.frame(groupTotals) %>%
  select(year, perp, tot) %>%
  group_by(year, perp) %>%
  summarize(tot = sum(tot)) %>%
  arrange(year) %>%
  spread(perp, tot) %>%
  dygraph() %>%
  dyLegend(show = "onmouseover") %>%
  dyHighlight(highlightSeriesBackgroundAlpha = .2,
              hideOnMouseOut = FALSE) %>%
  dyCSS("dygraphLegend.css")
```


## Another Example (Because They Are Fun!)

```{r, warning = FALSE, message = FALSE, echo = TRUE, eval = FALSE}
library(plotly)

plot_ly(economics, x = date, y = uempmed) %>%
  add_trace(y = fitted(loess(uempmed ~ as.numeric(date))), x = date) %>%
  layout(title = "Median duration of unemployment (in weeks)",
         showlegend = FALSE) %>%
  dplyr::filter(uempmed == max(uempmed)) %>%
  layout(annotations = list(x = date, y = uempmed, text = "Peak", showarrow = T))
```


## Another Example (Because They Are Fun!)

```{r, warning = FALSE, message = FALSE, echo = FALSE, eval = TRUE}
library(plotly)

plot_ly(economics, x = date, y = uempmed) %>%
  add_trace(y = fitted(loess(uempmed ~ as.numeric(date))), x = date) %>%
  layout(title = "Median duration of unemployment (in weeks)",
         showlegend = FALSE) %>%
  dplyr::filter(uempmed == max(uempmed)) %>%
  layout(annotations = list(x = date, y = uempmed, text = "Peak", showarrow = T))
```


## Last One...

```{r, warning = FALSE, message = FALSE, echo = TRUE, eval = FALSE}
library(ggplot2)

p = ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(text = paste("Transmission:", as.factor(am))), size = 2) +
  geom_smooth(aes(colour = as.ordered(cyl), fill = as.ordered(cyl)), 
              show.legend = FALSE) + 
  facet_grid(. ~ cyl) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  #scale_colour_discrete(name = "Cylinders") +
  lazerhawk::theme_trueMinimal()

ggplotly(p)
```



## Last One...

```{r, warning = FALSE, message = FALSE, echo = FALSE, eval = TRUE}
p = ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(text = paste("Transmission:", as.factor(am))), size = 2) +
  geom_smooth(aes(colour = as.ordered(cyl), fill = as.ordered(cyl)), show.legend = FALSE) + 
  facet_grid(. ~ cyl) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  #scale_colour_discrete(name = "Cylinders") +
  lazerhawk::theme_trueMinimal()

ggplotly(p)
```



## Interactive Tables

```{r, echo = 3, eval = TRUE}
mtcars1 = mtcars
mtcars1$am = factor(mtcars1$am, levels = 0:1)
DT::datatable(head(mtcars1), filter = "top")
```


## Document Generation

RStudio has built-in capacity to use <span class="pack">knitr</span> and <span class="pack">rmarkdown</span>.

These packages, in conjunction with bits and pieces of $\LaTeX$, allow you to create reproducible documents.

## Example

```{r, eval = FALSE, echo = TRUE}
`r lmSum = summary(lm(mpg ~ wt, data = mtcars))

if (lmSum$coefficients[2, 4] < .05) {
  paste("Weight's coefficient of", 
        round(lmSum$coefficients[2], 3), 
        "is significant", sep = " ")
} else {paste("Weight's coefficient of", 
        round(lmSum$coefficients[2], 3), 
        "is not significant", sep = " ")}`
```


```{r, echo = FALSE, eval = TRUE}
lmSum = summary(lm(mpg ~ wt, data = mtcars))

if (lmSum$coefficients[2, 4] < .05) {
  paste("Weight's coefficient of", 
        round(lmSum$coefficients[2], 3), 
        "is significant", sep = " ")
} else {paste("Weight's coefficient of", 
        round(lmSum$coefficients[2], 3), 
        "is not significant", sep = " ")}
```

## An Example In Practice

A good man once said:

> You, my dear sir, are but a mere
> `r paste(sample(c('artless','bawdy','beslubbering','bootless','churlish'), 1), sample(c('base-court','bat-fowling','beef-witted','beetle-headed','boil-brained'), 1), sample(c('apple-john','baggage','barnacle','bladder','boar-pig','bugbear'), 1))` 
> and I bid you a good day.

## The Code

```{r, eval = TRUE, echo = TRUE}
paste(sample(c('artless','bawdy','beslubbering','bootless'), 1), 
      sample(c('base-court','bat-fowling','beef-witted','beetle-headed'), 1),
      sample(c('apple-john','baggage','barnacle','bladder','boar-pig'), 1))
```

## Document Generation

Documents can be written out as PDF, HTML, or even Word.

The HTML documents can be pages or presentations.

With the exception of minor changes, it really does not take too much effort to switch between outputs.

There are also a handful of templates for you to use.

- All of the Tufte class documents are amazing.

## Projects And Built-in Version Control

RStudio let's you create a project with version control.

- Cloning from existing repositories

- Creating a new repository and "pushing" it up

- Version control can be through Git or Subversion.

## Package Management

There is nothing wrong with using the previously noted <span class="func">install.packages()</span> function.

RStudio has a menu-driven package installation function.

- It has a handy autocomplete function to it, so you can avoid any misspellings!

You are also given a list of all of your packages and which ones are currently loaded.

## Data Read

One of the most common ways to read data into R is through <span class="func">read.csv()</span>.

  - This will take care of almost everything you ever need to do.

Like all thing, RStudio tries to make things easier.

The most recent version of RStudio comes with functionality for <span class="pack">readr</span> and <span class="pack">haven</span>.

These packages allow for faster (and more sensible defaults) data reading from cvs, Excel, SAS, SPSS, and Stata. 

## Addins

Addins are relatively new to RStudio.

They are essentially functions that you can call interactively.

Especially useful if you find yourself creating the same text lines.

```{r, eval = FALSE, echo = TRUE}
lazerhawk::insertSlide()
```


# Wrap

## Wrapping It Up

Although our time was brief, I hope you can see the benefits of using R/RStudio for your research.

R only continues to develop, improve, and grow!

By using R, you will also find growth!

  - Maybe not happiness, but definitely growth.
  
```{r, eval = TRUE, echo = FALSE}
fortunes::fortune(which = 83)
```


