---
title: "Latent Variable Models"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: zenburn
    center: true
    transition: fade
    css: clean.css
    slide_level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, error = FALSE, message = FALSE, fig.align = "center")
library(DiagrammeR)
```


## The World Of Matrix Factorization

## Dimension Reduction

In a nutshell:

We have many variables.

We need to reduce the number of variables.

We extract components from the entire set of variables.

We want to be able to retain as much information from those components.

## PCA & FA

Principal Components Analysis (PCA) and Factor Analysis (FA) are frequently considered to be synonymous.

While they share a common goal in dimension reduction, there are some huge differences between them.

## PCA

Typically, the components from a PCA are restricted so they do not correlate with each.

PCA (shockingly) assumes no measurement error!

Scores (more on this later) don't mean what one might think.

SPSS users are likely aware that this is the default for the factor analysis menu.

The causal direction is a very important issue here.

## PCA -- Causal Flow

```{r, echo = FALSE}

grViz("
digraph dot {

graph [layout = dot, 
rankdir = BT]

node [shape = square,
      style = filled,
      color = paleturquoise]
pancake

node [shape = oval, 
      fillcolor = thistle]
eggs; milk; sugar; butter; flour

edge [color = grey]
eggs -> {pancake}
milk -> {pancake}
sugar -> {pancake}
butter -> {pancake}
flour -> {pancake}
}")

```


## FA

Factor rotations can be orthogonal or oblique.

The correlation matrix is constructed by including only variance common amongst the variables.

*Communality* is the sum of the squared loadings.

Scores reflect the underlying latent variable.

## FA -- Causal Flow

```{r, echo = FALSE}
grViz("
digraph dot {

graph [layout = dot]

node [shape = square,
      style = filled,
      color = paleturquoise]
pancake

node [shape = oval, 
      fillcolor = thistle]
eggs; milk; sugar; butter; flour

edge [color = grey]
pancake -> {eggs milk sugar butter flour}
}")

```

## Exploratory Factor Analysis 

## Number Of Factors

We all learned about the "elbow" and Kaiser.

```{r, echo = FALSE}
psych::scree(attitude)
```

## Doing Better

We can use *BIC* for model comparisons.

We can also use *parallel analysis* if we want to work in an *a prior* setting.

- PA does exhibit some sample size shenanigans.

## Reliability

Believe it or not, Cronbach's $\alpha$ is not the only measure of reliability!

It is incredibly easy to inflate and is driven by item number and correlation!

If you have anything other than a uni-dimensional factor structure, $\alpha$ is not an appropriate measure of reliability.

I might suggest McDonald's $\omega$. 

$\omega$ can take a few different forms.

$\omega_t$ provides a proportion of variance not unique to the factors.

$\omega_h$ is associated with the general factor in a hierarchical structure.

## $\omega_h$

```{r, echo = FALSE}
library(psych)

v9 <- sim.hierarchical()

myOmega = omega(v9)

omega.diagram(myOmega)
```


## Factor Scores

Frequently constructed as simple aggregates (e.g., sum, mean).

This assumes, however, that the variables have equal loadings.

And...they don't have any error.

They also do not reflect the latent variable.

They generally require recoding to happen.

## Factor Scores

Thurstone (regression) and Barlett scores are common.

Thurstone scores are pretty easy to produce:


$$ S = XW $$
$$ W = R^{-1}F $$

$X$ are the observed scores, $R$ is the correlation matrix, and $F$ is the factor loading matrix.


## Confirmatory Factor Analysis

Let us first jump over one big hurdle -- sample size!

Rules of thumb abound for all matters related to sample size.

Many estimators exist to deal with the issue of small, ill-behaved data.

They do not, however, mask the problem all together.

## Measurement Invariance

### Are comparisons possible?

Configural -- same model for each group

Weak -- equal unstandardized pattern coefficients

Strong -- equal unstandardized intercepts for indicators

Strict -- equal vcov



## Item Response Theory

Although PCA and FA exist outside of testing (they have rich traditions in Biology), we are still talking about *Classical Testing Theory*.

Item Response Theory offers a different take on "testing".

With IRT, we are examining the items and the person.

The main goal is to develop items that will *discriminate* between different *theta* levels.

## Flexible IRT

IRT is often (incorrectly) assumed to be for typical testing scenarios.

IRT is not just for dichotomous variables, but also polytomous variables.

IRT models can also contain different parameters (1PL, 2PL, 3PL, 4PL).

Error is not assumed to be consistent across individuals.

Models are not sample dependent.

Item difficulty and thetas can be compared.

## Item Characteristic Curves

Discrimination and Difficulty!

```{r, echo = FALSE}
library(mirt)

dat <- expand.table(LSAT7)

idealpt <- mirt(dat, 1, verbose = FALSE)

plot(idealpt, type = 'trace', facet_items = FALSE)
```


## Crawling Out Of The Continuum Minima

The latent variable world is much bigger than the "underlying continuum" world that we have been talking about.

Latent variables are everywhere and finding them is fun.

## Latent Dirichlet Allocation

LDA was developed by the superstars of the machine learning world.

The Dirichlet distribution is useful in many places.

But it is likely best known for one particular form of latent variable model...

## Topic Models!

While we can respect its place in a historical sense, content analysis is antiquated.

With topic models (and the LDA and MCMC that live in the weeds), we can do things with text that humans could never do.

For example...how long would it take you to read 10000 Amazon reviews?

## 10000 Amazon Reviews

```{r, echo = FALSE}
load("C:/Users/sberry5/Documents/class/lectureSlides/reviewTopicModel.RData")

library(stm)

plot.STM(reviewTM)
```


## Mixture Models

```{r, echo = FALSE}
library(ggplot2)

set.seed(1234)
erupt1 = rnorm(150, mean=2, sd=.25)
erupt2 = rnorm(150, mean=4.5, sd=.25)
mixed = sample(c(erupt1, erupt2))

ggplot() +
  geom_density(aes(x = mixed)) +
  theme_minimal()

```

## Clustering

```{r, echo = FALSE}
library(flexmix)
mod = flexmix(eruptions~1,  data=faithful, k = 2)

eruptionsNew = data.frame(faithful, cluster = as.factor(mod@cluster))

ggplot(eruptionsNew, aes(eruptions, waiting, color = cluster)) +
  geom_point() +
  theme_minimal()
```

## Wrapping Up

No matter what research world you currently live in, latent variables are everywhere.

Not only are they everywhere, but they are just waiting to be used.