---
title: |  
  | Regular Expressions
  | String Magic
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: zenburn
    center: true
    transition: fade
    css: clean.css
    slide_level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, error = FALSE)
library(dplyr)

```

## Why Am I Here?

Breakfast and magic!

Regular expressions (regex) can make you much more efficient.

Regex can also open new doors from an analysis perspective.


## When To Use Regex

String parsing and cleaning

String replacement (large scale)

You will never need to make changes manually again

### Not Just For Research

## How To Use Regex

It is like learning hieroglyphs, but much more useful.

You are just building patterns.

## {data-background-image="https://media.giphy.com/media/gMAQhQDwxl1fO/giphy.gif" data-background-size="contain"}

## Common Characters

```{r}
commonRegex = data.frame(regex = c("[a-zA-Z]", "[0-9]", "+", "^", "$"), 
                         use = c("Any letter", "Any number", "One or more", 
                                 "Start of string", "End of string"))

knitr::kable(commonRegex)
```

## Meta Characters

Some symbols mean certain things in regex.

```{r}
reservedChars = data.frame(meta = c('\\\\', "^", "$", 
                                       "{}", "\\[\\]", "()", 
                                       ".", "*", "+", "?", 
                                       "|", "<>", "-"), 
                           meaning = c("escape", "start", "end", "quantifier", 
                                       "one of contents", "capturing group", 
                                       "wildcard", "greedy zero or more", 
                                       "greedy one or more", "lazy quantifier", 
                                       "or", "start/end of word", "range"))

knitr::kable(reservedChars)
```

They all need to be escaped with a <span class="func">\\</span>

## {data-background-image="https://imgs.xkcd.com/comics/backslashes.png" data-background-size="contain"}


## A Simple Example

We have the following list of numbers:

```{r}
numList = c("618 549 5326", 
            "(618) 549-5326", 
            "618.549.5326", 
            "618-549-5326", 
            "6185495326", 
            "123654", 
            "357-84-5915")

numList
```


<a href="https://www.youtube.com/watch?v=AWU2GnPjlp0">Quatro's Deep Pan Pizza</a>

## Getting The Phone Numbers

```{r, results='markup'}
cat("^\\(*\\d{3}\\)*(\\s|-|.)*\\d{3}(\\s|-|.)*\\d{4}$")
```


```{r}
grep("^\\(*\\d{3}\\)*(\\s|-|.)*\\d{3}(\\s|-|.)*\\d{4}$", numList, value = TRUE)
```


## Broken Down

<span class="fragment highlight-blue">^\\(\*</span>
<span class="fragment highlight-red">\\d{3}</span>
<span class="fragment highlight-blue">\\)\*</span>
<span class="fragment highlight-green">(\\s|-|.)\*</span>
<span class="fragment highlight-red">\\d{3}</span>
<span class="fragment highlight-green">(\\s|-|.)\*</span>
<span class="fragment highlight-red">\\d{4}$</span>

## Globally Search a Regular Expression and Print

### Or grep to us regular folks

Using variants of grep let's us find our patterns.

## One Of These Things Is Not Like The Other...

Appropriate spacing

Common mistakes

Common mistakes 

 Common mistakes 


## The Problems
 
```{r}
spacing = c("Appropriate spacing", 
            "Common mistakes", 
            "Common mistakes ", 
            " Common mistakes", 
            " Common mistakes ", 
            "  Common mistakes")

spacing
```

## Isolating The Problems

```{r, results='markup'}
cat("^(\\s|\\t)+|(\\s|\\t)+$")
```

```{r}
grep("^(\\s|\\t)+|(\\s|\\t)+$", spacing, value = TRUE)
```


## Broken Down

<span class="fragment highlight-blue">^</span>
<span class="fragment highlight-red">(\\s</span>
<span class="fragment highlight-green">|</span>
<span class="fragment highlight-red">\\t)</span>
<span class="fragment highlight-blue">+</span>
<span class="fragment highlight-green">|</span>
<span class="fragment highlight-red">(\\s</span>
<span class="fragment highlight-green">|</span>
<span class="fragment highlight-red">\\t)</span>
<span class="fragment highlight-blue">+</span>
<span class="fragment highlight-blue">$</span>



## Where To Use Regex

Any decent programming language:

- JavaScript
- Python
- R

You can even use them in SAS and Stata.

## Less Obvious

### And Easier Than You Think

Google Docs

Notepad++ 

Find in Files

Qualtrics

Command line

Powershell

## Imagine If You Will...

A world where many documents (text, data, etc.) have problems.

- Little imagination required!

Without regular expressions, how would you solve this problem?

## {data-background-image="https://media.giphy.com/media/11xjcvOMD4HYFq/giphy.gif"}

## Find in Files

The *Find in Files* utility in Notepad++ can save you (or RAs) hours/days.

## 

We can see that just a little bit of regex can take us a long way.

Just as text can be elaborate, so can regular expressions.

## Infinite Complexity...

### Infinite Possibility

## Logic

## Boundaries

## POSIX Classes

## Lookarounds

## How To Learn

Dieter time!

## As Above, So Below

Once you start learning regular expressions, they start to come easier and finding patterns becomes almost second nature.