---
title: |  
  | Regular Expressions
  | String Magic
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: zenburn
    center: true
    transition: fade
    css: clean.css
    slide_level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE, error = FALSE)
library(dplyr)

```

## Why Am I Here?

To learn magic -- mostly!

In reality, regular expressions (regex) can make you much more efficient.


## When To Use Regex

### Not Just For Research

String parsing and cleaning

String replacement (large scale)

## How To Use Regex

Fortunately, we get to learn a new *language*.

- It is more akin to a series of symbols with definite meanings.

## Common Characters

```{r}
commonRegex = data.frame(regex = c("[a-zA-z]", "[0-9]", "+", "^", "$"), 
                         use = c("Any letter", "Any number", "One or more", 
                                 "Start of string", "End of string"))

knitr::kable(commonRegex)
```

## Meta Characters

Some symbols mean certain things in regex.

```{r}
reservedChars = data.frame(meta = c('\\\\', "^", "$", 
                                       "{}", "[]", "()", 
                                       ".", "*", "+", "?", 
                                       "|", "<>", "-"), 
                           meaning = c("escape", "start", "end", "quantifier", 
                                       "one of contents", "capturing group", 
                                       "wildcard", "greedy zero or more", 
                                       "greedy one or more", "lazy quantifier", 
                                       "or", "start/end of word", "range"))

knitr::kable(reservedChars)
```



## A Simple Example

We have the following list of numbers:

```{r}
numList = c("618 549 5326", 
            "(618) 549-5326", 
            "618.549.5326", 
            "618-549-5326", 
            "6185495326", 
            "123654", 
            "357-84-5915")

numList
```


Quatros Deep Pan Pizza

## Getting The Phone Numbers

```{r, results='markup'}
cat("^\\(*\\d{3}\\)*( |-|.)*\\d{3}( |-|.)*\\d{4}$")
```


```{r}
grep("^\\(*\\d{3}\\)*( |-|.)*\\d{3}( |-|.)*\\d{4}$", numList, value = TRUE)
```


## Where To Use Regex

Any decent programming language:

- JavaScript
- Python
- R

You can even use them in SAS and Stata.

## Less Obvious

### And Easier Than You Think

Google Docs

Notepad++ 

- Find in Files

Command line

Powershell

## Find in Files

The *Find in Files* utility in Notepad++ can save you (or RAs) hours/days.

## {data-background-image="https://media.giphy.com/media/11xjcvOMD4HYFq/giphy.gif"} 

## Imagine If You Will...

A world where many documents (text, data, etc.) have problems.

Little imagination required!



## Infinite Complexity...

### Infinite Possibility

Logic

Boundaries

POSIX Classes

Lookarounds

## How To Learn

## As Above, So Below