---
title: "Qualtrics: Beyond The Menus"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: zenburn
    center: true
    transition: fade
    css: clean.css
    slide_level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A Simple Request

If possible, please navigate to the following link: 

tinyurl.com/advQualDemo

## Cascading Style Sheets

CSS gives the web style.

Qualtrics gives you full control of the CSS for your surveys.

## Changing Skins -- Easy

You can click around in the Look & Feel menu.

## Changing Skins -- Powerful

Go to Look & Feel > Advanced > Add Custom CSS

## Changing Skins For Mobile

Buttons are radio buttons.

The Qualtrics 2014 works very well for mobile.

It is pretty bland, though, so spruce it some!

## With Great Power...

Having the ability to style everything is great.

It can help make your survey more accessible and branding always looks good.

You will need to know about:

<span class="fragment fade-up">web-based survey design</span>

<span class="fragment fade-up">human-computer interaction</span>

<span class="fragment fade-up">web standards</span>

<span class="fragment fade-up"><a href="https://nd.qualtrics.com/jfe/form/SV_2a7C6V4BhgqXs5n">Ye Have Been Warned!</a></span>

## JavaScript

JavaScript is a core technology for web content.

Try browsing the web with a script blocker!

It is responsible for everything from very basic sliders to some really advanced dynamic visualizations.

What is Qualtrics other than a series of pages?

## The Many Libraries of JS

JS has many different libraries:

<span class="func">jQuery.js</span> is everywhere (autocompletes anyone?)

<span class="func">Raphael.js</span> is great for creating svg

<span class="func">D3.js</span> powers interactive visualizations (NYT)

<span class="func">Prototype</span> 

<span class="func">Reveal.js</span>


## How To Use JS

There are a few different places you can put your JS.

Within the addOnLoad function in the JS editor.

In the JS editor without the addOnLoad function.

In the header.

In the question source.


## Why?

Flow control.

Creating new variables.

Fun times!

## Flow Control

Can be as simple as this:

```{r, eval = FALSE}
Qualtrics.SurveyEngine.addOnload(function() {
  
  this.questionclick = function(event, element) {
    if (this.getChoiceValue(1) == true) {
      this.clickPreviousButton()
    }
  }
  
})

```



## Flow Control Example

<a href="https://nd.qualtrics.com/SE/?SID=SV_3BN1Sxxz4XlKRLL">Glasses Example</a>


```{r, eval = FALSE}
Qualtrics.SurveyEngine.addOnload(function()
{
  $('NextButton') && $('NextButton').hide();	
  
  $('PreviousButton') && $('PreviousButton').hide();	
  
  var ros1 = document.getElementById("QR~QID2~1");
  var ros2 = document.getElementById("QR~QID2~2");
  var ros3 = document.getElementById("QR~QID2~3");
  
  ros1.disabled = true;
  ros2.disabled = true;
  ros3.disabled = true;
  
  function enableButton(){ros1.disabled = false;
  ros2.disabled = false;
  ros3.disabled = false;}
  
  setTimeout(enableButton, 10000)
  
  this.questionclick = function(event,element){
    if (element.type == 'radio'){
      this.clickNextButton();
    }
  }	
});
```


## New Variables

Using a combination of embedded data and JS enables us to do some interesting things.

## Good Times

We can also use JS to do some interesting things.


## { data-background-image="https://media.giphy.com/media/xpPLaEYLy637O/giphy.gif" data-background-size="400px" data-background-repeat="repeat"}


## Regular Expression

Regular expressions (regex) are used to match patterns in strings.

Think about "find & replace" and you are essentially there...

Although very powerful, regex is a bit confusing at first blush.

<span class="fragment fade-up"><span class="func">\\$[0-9]+.[0-9]{0, 2}</span></span>

<span class="fragment fade-up"><span class="func">\^[a-zA-Z]+\s.*[0-9]{4}\S$</span></span>

## Why Would I Need That?

Display Logic for questions, embedded data, panels, and IP.

Sometimes branching on certain text is helpful.

Survey Flow

Custom Validation

Reports

## Helpful Chunks


| Expression    |       Action    |
|:---------------:|:-----------------:|
| [0-9]         | Number range    |
| [a-z]         | Letter range    |
| ^             | String start    |
| $             | String end      |
| .             | Wild card       |
| *             | 0 or more times |
| +             | 1 or more times |
| \|            | Or operator     |
| ()            | Grouping        |
| {x,y}         | Specific matches|
| ?             | Ungreedy match  |


## Application Programming Interface

As an add-on (read: extra charge), Qualtrics provides an API.

Notre Dame was gracious enough to acquire this add-on.

## "API! I Don't Even Know What That Is!"

The most basic explanation is that APIs allow programs to talk to each other.

In addition to program-to-program, APIs also allow programs to chat with servers.

This is where the Qualtrics API becomes useful for us.

## Useful API Functions

https://api.qualtrics.com/

List Surveys

Import Survey

Create Response Export

## When To Use It

Using the API is probably the most difficult thing that we have discussed.

It really is only useful if you want to integrate it with your statistical programs.

Some programs make it easy (any OOPL), while others make it difficult/impossible.

There are two versions out there: V2 and V3.


## Why To Use It -- Efficiency

```{r, eval = FALSE}
exportQualtricsData = function (username, token, format, surveyID) {
  url = paste("http://survey.qualtrics.com//WRAPI/ControlPanel/api.php?Version=2.4&
              Request=getLegacyResponseData", 
              "&User=", userName, 
              "&Token=", userToken,
              "&Format=", format, 
              "&SurveyID=", surveyID, 
              "&ExportTags=1", 
              sep = "")
  
  url = gsub("[@]", "%40", url) %>% 
    gsub("[#]", "%23", .)
  
  exportQualtricsData = read.csv(url, stringsAsFactors = FALSE)
}

exampleData = exportQualtricsData("userName@nd.edu#nd", "userToken", 
                                  "CSV", "surveyID")
```


## { data-background-image="https://media.giphy.com/media/3oz8xOvhnSpVOs9xza/giphy.gif"}



## Wrapping Up

Right out of the box, Qualtrics can do just about anything you need.

Sometimes, though, we need to add features.

With a little bit of knowledge and fearlessness, we can pretty much do whatever we want.